Statically Typed String Sanitation Inside a Python: Technical Report

### 2014-11-17 sanitation-psp14/tr/strings_chapter.tex
Looks like a translation correctness proof
and some other minor changes.

% Judgements
\newcommand{\trden}[1]{\llbracket #1 \rrbracket} % = Translation Denotation.

% Relations
\newcommand{\treduces}{ \Downarrow }
\newcommand{\sreduces}{ \Downarrow }

@@

\title{Statically Typed String Sanitation Inside a Python: Technical Report}
\author{Nathan Fulton \\ Cyrus Omar \\ Jonathan Aldrich}

@@

@@

Theorems and lemmas appearing in \cite{fulton2014} are numbered, while
supporting facts appearing only in the Technical Report are lettered.

@@

@@

>>>>

% Judgements
\newcommand{\trden}[1]{\llbracket #1 \rrbracket} % = Translation Denotation.
% Relations
\newcommand{\treduces}{ \Downarrow }
\newcommand{\sreduces}{ \Downarrow }

@@

\title{Statically Typed String Sanitation Inside a Python: Technical Report}
\author{Nathan Fulton\and Cyrus Omar\and Jonathan Aldrich}

@@

Theorems and lemmas appearing in \cite{fulton2014} are numbered, while
supporting facts appearing only in the Technical Report are lettered.
Numbered items correspond to the numbering in \cite{fulton2014}.

@@

@@

\begin{thm}[Safety for $\lambdap$] If $\emptyset \vdash \iota : \tau$ 
  then $\iota \sreduces \dot v$ and $\emptyset \vdash \dot v : \tau$.
\end{thm}

We can also define canonical forms for regular expressions and strings in the usual way:

\begin{lem}[Canonical Forms for Target Language]
~
\begin{itemize}
\item 
If $\emptyset \vdash \iota : \regex$ then $\iota \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
\item 
If $\emptyset \vdash \iota : \str$ then $\iota \treduces \tstr{s}$.
\end{itemize}
\end{lem}

@@

\begin{thm}[Translation Correctness]\label{thm:trcorrect}
  If $\tctx \vdash e : \sigma$ then 
  there exists an $\iota$ such that $\trden{e} = \iota$
  and $\trden{\tctx} \vdash \iota : \trden{\sigma}$.
  Furthermore, $e \sreduces v$ and
  $\iota \treduces \dot{v}$ such that
  $\trden{v} = \dot{v}$.
\end{thm}
\begin{proof}
We present a proof by induction on the derivation that $\tctx \vdash e : \sigma$.
we write $e \leadsto \iota$ as shorthand for the final property.

\begin{itemize}[label=$ $, itemsep=1ex]
\item \textbf{Case $e = \rstr{s}$}. Suppose $\tctx \vdash \rstr{s} : \sigma$.

By examination the syntactic structure of conclusions in the relation S-T, 
we know this is true just in case $\sigma = \stringin{r}$ for some $r$ such that
$s \in \lang{r}$; and of course, there is always such an $r$.

There are no free variables in $\rstr{s}$, so we might as well proceed from the
fact that $\emptyset \vdash \rstr{s} : \stringin{r}$.

By definition of the translation ($\trden{\cdot}$) the following statements hold:
\begin{align}
\trden{\rstr{s}} &= \str{s} \\
\trden{\stringin{r}} &= \str \\
\trden{\emptyset} &= \emptyset
\end{align}
Note that $\emptyset \vdash \str{s} : \str$ by P-T-Str.
Recall that contexts are standard and, in particular, can be weakened.
So since $\trden{\tctx}$ is either a weakening of $\emptyset$ or $\emptyset$ itself,
$\trden{\tctx} \vdash \tstr{s} : \str$ by weakening.

Summarily, $\str{s}$ is a term of $\lambdap$ such that $\trden{\tctx} \vdash \str{s} : \trden{\sigma}$

It remains to be shown that there exist $v, \dot{v}$ such that $\rstr{s} \sreduces v$,
$\str{s} \treduces \dot{v}$, and $\trden{v} = \dot{v}$. But this is immediate because
each term evaluates to itself and we have already established the equality.

\item \textbf{Case $e = \rconcat{e_1}{e_2}$}. This case is an obvious appeal to induction.

\item \textbf{Case $e = \strcase{e_1}{e_2}{x,y.e_3}$}. This case relies on our definition
of context translation.

Suppose $\sctx \vdash \strcase{e_1}{e_2}{x,y.e_3} : \sigma$.
By inversion of the typing relation it follows that
$\sctx \vdash e_1 : \stringin{r}$, $\sctx \vdash e_2 : \sigma$ and $\sctx, x : \stringin{\lhead{r}}, y : \stringin{\ltail{r}} \vdash e_3 : \sigma$.

By induction, there exists an $\iota_1$ such that $\trden{e_1} = \iota_1$,
$\trden{\sctx} \vdash \iota_1 : \trden{\sigma}$, and $e_1 \leadsto \iota_1$.
Similarly for $e_2$ and some $\iota_2$.

By canonical forms, $e_1 \sreduces \rstr{s}$ and so $\iota_1 \treduces \tstr{s}$ by $\leadsto$.

Choose $\iota = \tconcat{\iota_1}{\iota_2}{x,y.\iota_3}$ \todo{and note that by the properties established via induction,
$\trden{e} = \iota$ and $\trden{\sctx} \vdash \iota : \trden{\sigma}$.}

Suppose $s = \epsilon$. Then $e \sreduces v$ where $e_2 \sreduces v$
and $\iota \treduces \dot{v}$ where $\iota_2 \treduces \dot{v}$.
But recall that $e_2 \leadsto v_2$ and so $\trden{v} = \dot{v}$.

Suppose otherwise that $s = at$ for some character $a$ and string $t$.
Then $e \sreduces v$ where $[a, t / x, y] e_3 \sreduces v$.
Similarly, $\iota \treduces \dot{v}$ where $[a, t / x, y] \iota_3 \treduces \dot{v}$

\begin{thm}[Correctness of Input Sanitation for Translated Terms]\label{thm:main}
  If $\trden{e} = \iota$ and $\emptyset \vdash e : \stringin{r}$ then $\iota \sreduces \tstr{s}$
  for $s \in \lang{r}$.
\end{thm}
\begin{proof}
  By Theorem \ref{thm:trcorrect} and the rules given, $\iota \sreduces \tstr{s}$ implies that $e \sreduces \strin{s}$.
  Theorem \ref{thm:scorrect} together with the assumption that $e$ is well-typed implies that $s \in \lang{r}$.
\end{proof}
